---
title: "AFCA_analysis"
output: html_document
date: "2025-04-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/projects/ageing_flies")
```

Load the libraries 
```{r}
library(anndata)
library(SeuratData)
library(Seurat)
library(SeuratDisk)
library(reticulate)
library(zellkonverter)
library(MAST)
library(tidyverse)
library(dplyr)
library(fs)
library(stats)
```

Set the file and folder info
```{r}
# Set the Results path
date <- format(Sys.Date(), "%Y%m%d")
results_path <- path("Results/age_scdgea", date)
if (!dir.exists(results_path)) {
  dir.create(results_path, recursive = TRUE)
}
```



Convert the Anndata to Seurat
```{r}
file.exists("Data/adata_body_filtered_2.h5ad")
ad <- readH5AD("Data/adata_body_filtered_2.h5ad")
adata_Seurat <- as.Seurat(ad, counts = "X", data = NULL)
rm(ad)
```

Convert the needed variables to factors 
```{r}
adata_Seurat$age <- as.factor(adata_Seurat$age)
adata_Seurat$batch <- as.factor(adata_Seurat$batch)
adata_Seurat$sex_age <- as.factor(adata_Seurat$sex_age)
adata_Seurat$afca_annotation <- as.factor(adata_Seurat$afca_annotation)
```


Define the groups and set the output file path
```{r}
# Define groups
young_group <- "5"
all_groups <- unique(adata_Seurat$age)
other_groups <- setdiff(all_groups, young_group)

# Get unique cell types (as character vector)
cell_types <- unique(as.character(adata_Seurat$afca_annotation))
print(cell_types)

```

Perform differential expression analysis per cell type
```{r}
# Set output file path
results_file <- file.path(results_path, "combined_markers.csv")
# remove file if it exists
if (file.exists(results_file)) {
  file.remove(results_file)
}

# Begin looping over cell types and comparisons
for (ct in cell_types) {
  cell_subset <-  subset(adata_Seurat, subset = afca_annotation == as.character(ct))
  cat(paste("\n Covering the cell type:", ct, "\n"))

  groups_present <- unique(cell_subset$age)

  if (young_group %in% groups_present) {
    for (grp in other_groups) {
      if (grp %in% groups_present) {
        cat(paste("  Running DGEA for:", young_group, "vs", grp, "\n"))
        try({
          markers <- FindMarkers(cell_subset,
                                 ident.1 = young_group,
                                 ident.2 = grp,
                                 group.by = "age",
                                 test.use = "wilcox",
                                 # latent.vars = c("batch", "sex")
                                 )

          markers$cell_type <- ct
          markers$comparison <- paste(young_group, "vs", grp)
          markers$gene <- rownames(markers)
          # rename the BF corretion
          markers <- markers %>% rename(
            p_val_BF = p_val_adj
            )
          # perform the BH correction
          markers$p_val_adj <- p.adjust(markers$p_val, method = "BH")

          # Write to CSV file (with header if first write)
          write.table(markers, results_file,
                      sep = ",",
                      row.names = FALSE,
                      col.names = !file.exists(results_file),
                      append = file.exists(results_file))
        }, silent = TRUE)
      }
    }
  }
}

# # update the df by adjusting the pvalues with BH correction
# combined_markers <- read.csv( "Results/combined_markers.csv") %>%
#   group_by(cell_type) %>%
#   mutate(adj_p_val_within_cell_type = p.adjust(p_val, method = "BH")) %>%
#   ungroup()
# 
# # Save the final version of the df
# write.csv(combined_markers, "Results/full_combined_markers.csv", row.names = FALSE)


cat("All results saved incrementally to:\n", results_file, "\n")

```


Refine the df for more measured analysis
```{r}
# Create list with only the enterocyte types
enterocyte_types <- unique(adata_Seurat$afca_annotation) %>% 
  str_subset("enterocyte")

# Create a new df with only the enterocyte types
enterocyte_df <- adata_Seurat %>% 
  subset(afca_annotation %in% enterocyte_types)
```

Perform differential expression analysis between age groups without subsetting by cell type
```{r}
# Set output file path
results_file <- file.path(results_path, "entero_markers.csv")
# remove file if it exists
if (file.exists(results_file)) {
  file.remove(results_file)
}

# loop over comparisons
groups_present <- unique(enterocyte_df$age)
for (grp in other_groups) {
  if (grp %in% groups_present) {
    cat(paste("  Running DGEA for:", young_group, "vs", grp, "\n"))
    try({
      markers <- FindMarkers(enterocyte_df,
                             ident.1 = young_group,
                             ident.2 = grp,
                             group.by = "age",
                             test.use = "wilcox",
                             # latent.vars = c("batch", "sex") ONLY FOR MAST
                             )
      
      # add cell type and comparison columns
      markers$comparison <- paste(young_group, "vs", grp)
      markers$gene <- rownames(markers)
      # rename the BF corretion
      markers <- markers %>% rename(
        p_val_BF = p_val_adj
      )
      # perform the BH correction
      markers$p_val_adj <- p.adjust(markers$p_val, method = "BH")
  
      # Write to CSV file (with header if first write)
      write.table(markers, results_file,
                  sep = ",",
                  row.names = FALSE,
                  col.names = !file.exists(results_file),
                  append = file.exists(results_file))
    }, silent = TRUE)
  }
}

```




INTERPRETATION & SUMMARISATION

Load in the completed data and set paths
```{r}
diff_exp <- read.csv("Results/full_combined_markers.csv")
adata_meta <- adata_Seurat@meta.data
adata_meta$age <- as.integer(as.character(adata_meta$age))
rm(adata_Seurat)
```

Summarise the expression data
```{r}
dif_exp_sum <- diff_exp %>%
  group_by(cell_type, comparison) %>% 
  summarise(
    n_signif =  sum(adj_p_val_2 <= 0.1),
    n_insignif = sum(adj_p_val_2 > 0.1),
    total_genes = n(),
    signif_ratio = round(
      sum(adj_p_val_2 <= 0.1)/sum(adj_p_val_2>0.1),
      5)
  ) %>%
  mutate(
    age_1 = as.integer(str_split(comparison, 
                      " ")[[1]][1]),
    age_2= as.integer(str_split(comparison, 
                      " ")[[1]][3])
  )
```



Explore the samples from the original data

Cell type summary to show per cell type: 
- number of samples (flies)
- number of cells
- ratio of male/female
- average n_genes_by_count
- average total count


```{r}
ct_summary <- adata_meta %>%
    group_by(afca_annotation, age) %>% 
    summarise(
      cell_count = n(),
      sample_count = n_distinct(batch),
      sex_ratio = round(sum(sex == "male")/sum(sex == "female"), 2),
      avg_n_genes_by_cnt = round(mean(n_genes_by_counts),2),
      avg_ttl_cnts = round(mean(total_counts),2),
      avg_mt_genes = round(mean(total_counts_mt),2), 
      avg_mt_pct = round(mean(pct_counts_mt),2)
      )
    
```

Age comparison summary to show per cell type: 
- ratio of samples (flies)
- ratio of cells
- ratio of sex ratio
- ratio of average n_genes_by_count
- ratio average total count
```{r}
comp_summary <- ct_summary %>% 
  select(-c("avg_mt_genes", "avg_mt_pct")) %>% 
  mutate(
    "5 vs 30" = ifelse((age == 5|age == 30), 1, 0),
    "5 vs 50" = ifelse((age == 5|age == 50), 1, 0),
    "5 vs 70" = ifelse((age == 5|age == 70), 1, 0)
  ) %>%
  pivot_longer(
    cols = c("5v30", "5v50", "5v70"),
    names_to = "comparison",
    values_to = "included"
  ) %>%
  filter(included == 1) %>%
  select(-included) %>%
  pivot_wider(
    id_cols = c(afca_annotation, comparison),
    names_from = age,
    values_from = c(cell_count, sample_count, sex_ratio,
                    avg_n_genes_by_cnt, avg_ttl_cnts),
    names_sep = "_age_"
  ) %>%
  mutate(
    sample_comp = paste0(sample_count_age_5, ":", coalesce(
      sample_count_age_30, 
      sample_count_age_50, 
      sample_count_age_70
    )),
    cells_comp = paste0(cell_count_age_5, ":", coalesce(
      cell_count_age_30, 
      cell_count_age_50, 
      cell_count_age_70
    )),
    ratio_samples = sample_count_age_5 / coalesce(
      sample_count_age_30, 
      sample_count_age_50, 
      sample_count_age_70
    ),
    ratio_cells = cell_count_age_5 / coalesce(
      cell_count_age_30, 
      cell_count_age_50, 
      cell_count_age_70
    ),
    ratio_sex = sex_ratio_age_5 / coalesce(
      sex_ratio_age_30, 
      sex_ratio_age_50, 
      sex_ratio_age_70
    ),
    ratio_avg_genes = avg_n_genes_by_cnt_age_5 / coalesce(
      avg_n_genes_by_cnt_age_30, 
      avg_n_genes_by_cnt_age_50, 
      avg_n_genes_by_cnt_age_70
    ),
    ratio_avg_counts = avg_ttl_cnts_age_5 / coalesce(
      avg_ttl_cnts_age_30, 
      avg_ttl_cnts_age_50, 
      avg_ttl_cnts_age_70
    )
  ) %>%
  mutate(across(starts_with("ratio_"), ~round(., 2))) %>%
  select(c("comparison","sample_comp", "cells_comp", starts_with("ratio"))) %>% 
  left_join(
    ., dif_exp_sum, by = join_by("afca_annotation" == "cell_type", "")
  )
    
```











